$Z_3$ topological order in the quantum dimer-pentamer model

### 2016-08-06 qdpm_paper.tex
fig updates

While on non-bipartite lattices, the QDM posses a topologically ordered ground state~\cite{Moessner2001a,Fendley2002}, on the square lattice there is only a liquid ground state at an isolated critical quantum point~\cite{Leung1996,Syljuasen2006}. If the hard-core constraint of the QDM is relaxed to a parity constraint such that each vertex much have an odd (or even) number of dimers touching each vertex, the simplest non-trivial dynamics leads to a $Z_2$ topologically ordered ground state~\cite{Kitaev2003,Wen2003}. The existence of topologically ordered ground states in dimer-based models is intimately related to the local constraint which imposes a local gauge symmetry in the Hilbert space; the topological order that arises is directly related to a discrete gauge theory with the appropriate local gauge constraint~\cite{Moessner2001}.

Given that these models with dimer degrees of freedom and local constraints posses the simplest non-trivial local (Ising) Hilbert space at each link, they provide a potentially simpler route to engineering exotic phases. The local constraints can often be enforced by a local potential energy penalty and the required dynamics are generally expected to arise at the lowest order non-trivial order in perturbation theory. We will refer to such models with dimer degrees of freedom on links and local constraints at vertices as \emph{generalized quantum dimer modes} (GQDMs). Previous work as demonstrated that GQDMs on the square lattice support gapped $Z_2$ topologically ordered phases as well as gapless quantum critical points. On non-bipartite lattices, GQDMs support $Z_2$ topological order~\cite{Moessner2001a,Misguich2002} as well as doubled-semion phases~\cite{Qi2014,Buerschaper2014a}. A full characterization of all exotic phases that can exist within this framework is thus desirable.

@@

This this paper is organized as follows. In \secref{sec:Background} we provide relevant background for the QDM and related models. In \secref{sec:QDPM} we present the details of the QPDM. \secref{sec:MC} provides the results of a Monte Carlo study of the ground state of the QDPM at its exactly soluble point.

@@

%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[]
        \centering
        \includegraphics[width=1.0\columnwidth]{monomer_cor_log.pdf}
        \caption{ The monomer-monomer correlations function $\langle m_{\bm{0}} m_{\bm{r}} \rangle$ along a coordinate axis in the QDPM RK ground state doped with a single pair of monomers on lattices up to \note{$L=48$}. The horizontal lines correspond to $L^{-4}$ for the system size corresponding to the color. \note{Owen: Do you have data for $L=64$? Can you make the subscripts in the y-axis label bold math? Maybe we should renomalize this plot by a factor of $L^2$. This is only one sub-lattice--what about the other sublattice?}}
        \label{fig:monomer_correlations}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
    \centering
    \includegraphics[width=1.0\columnwidth]{spatial_cors_perp.pdf}
    \caption{Correlation function between perpendicular dimers on the same (blue) and different (red) 
    sub-lattice for a $L=64$ lattice.}
    \label{fig:spatial_dmr_cor}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

>>>>

While on non-bipartite lattices, the QDM posses a topologically ordered ground state~\cite{Moessner2001a,Fendley2002}, on the square lattice there is only a liquid ground state at an isolated critical quantum point~\cite{Leung1996,Syljuasen2006}. If the hard-core constraint of the QDM is relaxed to a parity constraint such that each vertex must have an odd (or even) number of dimers touching each vertex, the simplest non-trivial dynamics leads to a $Z_2$ topologically ordered ground state~\cite{Kitaev2003,Wen2003}. The existence of topologically ordered ground states in dimer-based models is intimately related to the local constraint which imposes a local gauge symmetry in the Hilbert space; the topological order that arises is directly related to a discrete gauge theory with the appropriate local gauge constraint~\cite{Moessner2001}.

Given that these models with dimer degrees of freedom and local constraints posses the simplest non-trivial local (Ising) Hilbert space at each link, they provide a potentially simpler route to engineering exotic phases. The local constraints can often be enforced by a local potential energy penalty and the required dynamics are generally expected to arise at the lowest order non-trivial order in perturbation theory. We will refer to such models with dimer degrees of freedom on links and local constraints at vertices as \emph{generalized quantum dimer modes} (GQDMs). Previous work has demonstrated that GQDMs on the square lattice support gapped $Z_2$ topologically ordered phases as well as gapless quantum critical points. On non-bipartite lattices, GQDMs support $Z_2$ topological order~\cite{Moessner2001a,Misguich2002} as well as doubled-semion phases~\cite{Qi2014,Buerschaper2014a}. A full characterization of all exotic phases that can exist within this framework is thus desirable.

@@

This paper is organized as follows. In \secref{sec:Background} we provide relevant background for the QDM and related models. In \secref{sec:QDPM} we present the details of the QPDM. \secref{sec:MC} provides the results of a Monte Carlo study of the ground state of the QDPM at its exactly soluble point.

@@

%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Monomer correlations}

%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[]
        \centering
        \includegraphics[width=1.0\columnwidth]{monomer_cor_log.pdf}
        \caption{ The monomer-monomer correlations function $\langle m_{\bm{0}} m_{\bm{r}} \rangle$ along a coordinate axis in the QDPM RK ground state doped with a single pair of monomers on lattices up to \note{$L=48$}. The horizontal lines correspond to $L^{-4}$ for the system size corresponding to the color. \note{Owen: Do you have data for $L=64$? Can you make the subscripts in the y-axis label bold math? Maybe we should renomalize this plot by a factor of $L^2$. This is only one sub-lattice--what about the other sublattice?}}
        \label{fig:monomer_correlations}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
    \begin{tabular}{| l | c | c |}
  	\hline
     & $c$ & $\bar{c}$ \\
	\hline
    $i$ & $\monomermvA$ &  $\monomermvB$  \\
    \hline
  \end{tabular}
  \caption{The pair of local dimer \& monomer configurations $c\leftrightarrow\bar{c}$ generate
  minimal local monomer dynamics. Monomers always remain on their origonal sublattice. }
  \label{tab:pentamerMoves}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
    \centering
    \includegraphics[width=1.0\columnwidth]{spatial_cors_end_to_end_parallel.pdf}
    \caption{Correlation function between parallel dimers (end-to-end) on the same (blue) and different (red) 
    sub-lattice for a $L=64$ lattice.}
    \label{fig:spatial_dmr_cor}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
